---
title: "01 - Interest Calculations"
author: "Michael Tychonievich, Ph.D."
output: html_notebook
---

Let's make some more scripts to deal with interest rate problems.  To get the
most out of this notebook, you should open up a separate .r script file and
retype the code there.  Then, when it comes time to change the code or build on
it, you should save a backup copy and continue your work.  Leave enough comments
behind that you will know what you were trying to do if you look back at your
work later.

### Nominal and effective interest rates

1.  Make a calculator that, given a nominal interest rate per year, a number 
of compounding periods per year, and a second number of compoundings per year,
returns the equivalent nominal interest rate.

Solution.  I started with the formula that converts nominal interest rate per
year into an accumulation factor for compound interest.  Since the interest rate
is constant, it's enough to do this for one year.  I wrote the equation down
again and did some algebra to solve for the new nominal rate.  Then I typed in
the calculation in three steps to help with readability.

```{r}
new_nominal <- function(old_nominal, old_periods, new_periods) {
  eff_ann_acc_factor <- (1 + old_nominal/old_periods)^old_periods
  new_acc_factor_per <- (eff_ann_acc_factor)^(1/new_periods)
  new_nominal_rate <- new_periods * (new_acc_factor_per - 1)
  return(new_nominal_rate)
}
```

2.  Write a script that makes a table to verify this calculation numerically.

Solution.  We've already for the accumulation factor for the first setup.  Let's
make a function to find it for the second setup and compare the results for a
few different sets of inputs.  I produced these inputs uniformly at random using
the runif() function, which is a way to get test data automatically.  The table
I made is quite ugly and may not present the information the way I want, but 
does it verify the calculation?  How would you improve its appearance and the
quality of the information it delivers? 

```{r}
new_nominal <- function(old_nominal, old_periods, new_periods) {
  eff_ann_acc_factor <- (1 + old_nominal/old_periods)^old_periods
  new_acc_factor_per <- (eff_ann_acc_factor)^(1/new_periods)
  new_nominal_rate <- new_periods * (new_acc_factor_per - 1)
  return(new_nominal_rate)
}

make_row <- function(old_nominal, old_periods, new_periods) {
  old_ann_acc_factor <- (1 + old_nominal/old_periods)^old_periods
  new_nominal_rate <- new_nominal(old_nominal, old_periods, new_periods)
  new_ann_acc_factor <- (1 + new_nominal_rate/new_periods)^new_periods
  row = c(old_ann_acc_factor, new_ann_acc_factor, 
          old_ann_acc_factor - new_ann_acc_factor)
  return(row)
}

make_table <- function() {
  comparison_table <- c("old acc", "new acc", "residual: old - new")
  for (i in 1:20) {
    new_row <- make_row(runif(1,0,1), runif(1,1,100), runif(1,1,100))
    comparison_table <- rbind(comparison_table, new_row)
  }
  return(comparison_table)
}
make_table()
```

3.  "Automatic testing."  Make a function that automates the testing of the 
function new_nominal, compiling data into some sort of histogram for residuals.

Solution.  I will take some of the code from above and reuse it.  In particular,
I want to extract the part that generates random data and produces residuals.
I decided to redo that part of the code this time to take advantage of a
feature that R has: if you have a for loop that only repeatedly applies the same
function over a large data set, you can use a function called mapply() instead
of writing out the full for loop.  This makes the code more readable (once you
know what mapply() is for) and is common in "R-style" coding.

What do you think of the output histogram?  What happens if you increase the
number of trials?  What happens if you allow for larger numbers of periods?

```{r}
new_nominal <- function(old_nominal, old_periods, new_periods) {
  eff_ann_acc_factor <- (1 + old_nominal/old_periods)^old_periods
  new_acc_factor_per <- (eff_ann_acc_factor)^(1/new_periods)
  new_nominal_rate <- new_periods * (new_acc_factor_per - 1)
  return(new_nominal_rate)
}

make_test_data <- function(num_trials=1000) {
  old_nominal <- runif(num_trials, 0, 1)
  old_periods <- runif(num_trials, 1, 100)
  new_periods <- runif(num_trials, 1, 100)
  #                   mapply(my_function, my_data)
  new_nominal_rate <- mapply(new_nominal, old_nominal, old_periods, new_periods)
  old_ann_acc_factor <- (1 + old_nominal/old_periods)^old_periods
  new_ann_acc_factor <- (1 + new_nominal_rate/new_periods)^new_periods
  return(old_ann_acc_factor - new_ann_acc_factor)
}

compile_test_data <- function(num_trials=1000) {
  trials <- make_test_data(num_trials)
  residuals_hist = matrix(0, nrow=4, ncol=1)
  rownames(residuals_hist) <- c(">10^-13", "<10^-13", "<10^-14", "<10^-15")
  colnames(residuals_hist) <- "number of trials"
  for (i in 1:num_trials) {
    if (trials[i] < 10^-15) {
      residuals_hist["<10^-15",] = residuals_hist["<10^-15",] + 1
    } else if (trials[i] < 10^-14) {
      residuals_hist["<10^-14",] = residuals_hist["<10^-14",] + 1
    } else if (trials[i] < 10^-13) {
      residuals_hist["<10^-13",] = residuals_hist["<10^-13",] + 1
    } else {
      residuals_hist[">10^-13",] = residuals_hist[">10^-13",] + 1
    }
  }
  return(residuals_hist)
}

compile_test_data(1000)
```

4.  Rewrite make_test_data() to replace the use of mapply() with a for loop.

5.  Rewrite the scripts from 1. and 2. using "mathematical" notation.  Does this
improve readability?

6.  Repeat exercises 1 and 2, only this time you are given the new nominal
interest rate and your goal will be to calculate the new number of compoundings
per year.  Will it always be mathematically possible for your calculator to give
an answer?

7.  Challenge: use kwargs to combine the two calculators into a
single function.

### Present value










