---
title: "01 - Calculation"
author: Michael Tychonievich, Ph.D.
output: html_notebook
---

### Calculator operations

To get you used to using R, I will start by showing you how to use R to turn
your computer into a fancy calculator.  We'll use RStudio's notebook feature.
To run code in the notebook, you can click the green arrow in the upper right
hand corner of the code cell or press Ctrl+Shift+Enter on your keyboard.  Let's
start with some simple arithmetic:

```{r}
2 + 2
3 + 5
5 / 7
5 ^ 4
9 + 7
2*pi / 6
```

Notice how the outputs come out in the order that they were entered into the
console.  You can ask R to evaluate very complicated expressions if you like;
use parentheses to force your preferred order of operations.

```{r}
((2 + 5/8)^(2/3))*9
```

There are some other arithmetic operations that you will want to remember:

#### _a %% b_: residue of _a_ modulo _b_

This is the remainder you get when you divide _a_ by _b_, taken as a number
between _0_ and _b_.  The inputs and outputs to this do not have to be integers,
but the **quotient** (that you don't see) does.  This operation will be very
handy when we want to have a program model events that happen with some regular
frequency that is not "every time period."  In math, this residue is always
positive, so be careful when moving from one subject to another.

```{r}
10 %% 3
3 %% 5
10 %% pi
10 %% -pi
```

#### _a %/% b_: integer quotient of _a_ divided by _b_

This divides _a_ by _b_, leaving the result as an integer.  It takes the largest
integer _q_ so that $q \cdot b < a$ as long as _b_ is positive, and it does
something similar if _b_ is negative.

```{r}
10 %/% 3
10 %/% -3
2.2 %/% 1.2
```

There is a relationship between _%%_ and _%/%_ called quotient-remainder that
you should remember: _a = b*(a %/% b) + (a %% b)_.  Let's try it out:

```{r}
3*(10 %/% 3) + (10 %% 3)
(-3)*(10 %/% -3) + (10 %% -3)
pi*(4.3 %/% pi) + (4.3 %% pi)
```


### Variables

You'll often want to store a number or other object for later use.  R does this
by assigning a value to a variable.  There are two different ways to do this in
R, as follows:

```{r}
x = 57
x
y <- 33
y
```

The preferred way to make assignments in R is with the arrow operator (<-), but
the equality symbol _usually_ does the same thing.

By default, R will not print any output to the console when you make a variable
assignment, you will need to do something if you want to see the number that
you you just saved a reference to.  Here are three ways to make this happen:

```{r}
x = 57
print(x)
y <- 33
y
(z <- 31)
```

When assigning values to variables, you can make references to other variables
that you already assigned.  You'll see these variables and a summary of their
values in your "Environment" window.

```{r}
x <- 57
y <- 33
z <- x + y
print(z)

```

It is completely fine to go back and change the value of a variable that you
previously defined, but realize that this will not change the values of
any variables that you used the first variable to define:

```{r}
x <- 57
y <- 33
z <- x + y
print(z)
x <- 0
print(z)

```

I useful consequence of this is that you can redefine a variable in a way that
uses its current value.  When this happens, the old value is effectively
removed and a new value replaces it.  _It is better to think of variables as
names for storage sites for your data rather than as names for the data
themselves_.

```{r}
x <- 57
print(x)
x <- x^2
print(x)

```

### Built-in functions

As with a graphing calculator, R comes with a large number of useful
mathematical functions already built in.  It would be unreasonable to list them
here, so I will go over some of the most important examples

#### Logarithms

To access logarithm functions, you may use the _log()_ function.  In R, the
default base for _log()_ is _e_.  If you want to use another base, you must
explicitly say so through the use of what is known as a keyword argument, or
"kwarg".  To
specify a keyword argument for the keyword "base" that determines the base of
the logarithm, you must include a statement that begins with "base = " on your
list of inputs to the function.  Since _log()_ only takes one keyword argument,
it is actually okay to omit the explicit mention of the "base" input, but it is
good practice for dealing with more complicated functions that may involve
multiple keyword arguements, only a few of which you want to actually change
from the default values.  Which of the ways of calling for the base-10 logarithm
is easier for you to read?

```{r}
log(10)
log(10, base = exp(1))
log(10, base = 10)
log(10, 10)

```

#### Square root

To take a square root, either use ^ or the _sqrt()_ function.  There is no
guarantee that these will calculate the square root in the same way, but that
shouldn't be an issue.

```{r}
sqrt(5)
5^(0.5)
5^(1/2)

```

#### Rounding

There are many ways of taking a number and getting a "nearby" integer in R.
Here are a few examples:

```{r}
x = 3.4
round(x)
ceiling(x)
floor(x)
trunc(x)
as.integer(x)
```

Note that with computers, numerical values involving decimals are almost never
recorded exactly.  This means that these operations may not always do exactly
what you expected when it comes to input values near where the functions change
values abruptly.  Both of the following two numbers should both be just slightly
larger than 1, and so _ceiling()_ should give value 2 in both cases.

```{r}
ceiling(1 + 1*10^(-15))
ceiling(1 + 1*10^(-16))

```
# Exercises.

1.  Let's pull up the curtain and take a look at how R assigned variables.  Here
I will use a potentially dangerous function (.Internal) to ask R to tell me
where in my computer the values I made are physically located.  The information
I am interested in is that first block of characters, starting with "@".  What
can you say about what you see?

```{r}
print("Assign x and y:")
x <- 2
y <- x
.Internal(inspect(x))
.Internal(inspect(y))
print("Reassign y:")
y <- 2
.Internal(inspect(x))
.Internal(inspect(y))
print("Reassign x and y:")
x <- 4
y <- 4
.Internal(inspect(x))
.Internal(inspect(y))
```

2.  In the section on rounding, you saw the following script.  By changing the
value of x, try to figure out what each of those rounding functions actually
does.  Use ?round to look at the help file and see if you figured their behavior
out completely.

```{r}
x = 3.4
round(x)
ceiling(x)
floor(x)
trunc(x)
as.integer(x)
```