---
title: "03 - Vectors"
author: Michael Tychonievich, Ph.D.
output: html_notebook
---

### Vectors: R's default kind of "list"

It is often useful to store series of numbers in a list so that they can all be
processed and moved together, using a nice naming scheme.  The default way that
R does this is by forming a vector.  **CAUTION: All numbers in a single **
**vector must be of the same type!**  If you form a vector with different data
types, R will coerce
all of the data according to the rule given in the previous notebook.  This
usually won't be a problem, but beware placing a character string in a vector
with numbers!  The simplest way to form a vector from numbers that you happen
to have lying around is to use the _c()_ function.

```{r}
x <- c(1, 5, 7)
x
y <- c("a", 5, 7, 3)
y
```

Note that the numerical values gained quotation marks in the second vector.
This indicates that they are no considered to be character strings by R and
not numerical values.  R does this so that each vector has a single format
associated with it, meaning that it can save on memory by not having to remember
the format of each entry individually.  It also means that R does not have to
think as much about how to combine two vectors together using arithmetic
operations.  This matters a lot when dealing with
very large data sets, even if it is somewhat of a hassle now.  

If you really want to get a list
of values with different types, use the _list()_ function as you would use the
_c()_ function.  Lists are more flexible in terms of storing data, but none of
arithmetic operations below work easily with them.  I will try to avoid using
lists whenever possible.

### Vector arithmetic

Vectors can be combined in R in several ways.  The general rule to remember is
that R will combine vectors entry-by-entry unless it can't or you go out of your
way to ask for something else.

```{r}
x <- c(1, 5, 7)
y <- c(2, 1, 5)
x + y
x * y
x ^ y
```

When you perform arithmetic operations involving vectors of different lengths,
R will do its best to match the sizes and then perform the operation.  This
means that the smaller vector will be combined to itself until it is as long
as the other vector, and then the operation will take place.

```{r}
x <- c(1, 5, 7)
z <- c(2, 1, 5, 6, 3, 9, 1, 1, 1)
x * z
z * x
```

In particular, this works well with scalar addition and multiplication, as long
as you remember that adding a single number to a vector adds it to every entry
of the vector separately.

```{r}
x <- c(1, 5, 7)
a <- 2
x + a
x * a
```

If the length of the smaller vector does not divide the length of the longer
vector evenly, you will get an error message.  If you accidentally combine
vectors of different lengths, it is important to remember that sometimes R will
throw back an error, and sometimes R will act by making compies of the smaller
vector and then combine the resulting vectors as you asked!  You can go ahead
and check the length of the vector with the _length()_ function if you like.

```{r}
x <- c(1, 5, 7)
z <- c(2, 1, 5, 6, 3, 9, 1, 1)
length(x)
length(z)
```

### Combining vectors

If you want to combine two vectors into a single longer vector, the _c()_
function handles that as well.  Notice that the order of the vectors in the
argument list matters.

```{r}
x <- c(1, 5, 7)
z <- c(2, 1, 5, 6, 3, 9, 1, 1)
c(z, x)
c(x, z)
```

### Vector indices

The _c()_ function works great when it comes to joining two vectors together,
but it obviously cannot handle pulling out pieces of a longer vector.  R has
a completely different syntax for this, using indices.  In a vector, each of the
various entries each get an integer index (a name or label) associated with it.
The first element of the vector gets the index 1, and the index goes up by one
for each entry you move to the right.  You can access the entry with index _i_
of the vector _z_ by typing _z[i]_.

```{r}
z <- c(2, 1, 5, 6, 3, 9, 1, 1)
z[1]
z[2]
c(z[3], z[5])
z[9]
```

For that last entry, R gave back a new sort of value, NA.  This value indicates
that R was unable to find any data at that index, but it leaves open the
possibility that the number exists while R just doesn't know what the number's
value is.  Think of it as R's indicator for "missing" data.  This is not the
same as R's NULL object, which indicates that data is actually nonexistent.

We can change the entries of a vector using this notation as well.  Doing so
will have a similar affect to reassigning the vector, in that it won't change
any data that you already computed from the old version of the vector.  On the
other hand, it involves less use of computer resources than completely
remaking the vector, which can matter when dealing with large data sets and
multi-stop algorithms.  You can even change a value to NA!

```{r}
z <- c(2, 1, 5, 6, 3, 9, 1, 1); z
z[1] <- 3; z
x <- c(1, 5, 7); x
x[2] <- NA; x
```

### Vector "slicing"

If you want to grab a subset of the entries of a vector, it is possible to
get them by using indices explicitly.  Since this sort of operation is so
commonly used, R has some shortcuts in place to take "slices" out of a vector.
You do this by merely listing the indices of the entries that you want, in the
order that you want them.

```{r}
z <- c(2, 1, 5, 6, 3, 9, 1, 1); z
z[c(1, 2, 3)]
z[c(4, 8)]
z[c(9, 8, 7, 6)]
```

For certain commonly requested slices, R has syntax to make the requests even
faster.  The most useful of these is the colon function, which generates lists
of numbers in arithmetic progression.  Here are some examples:

```{r}
1:10
10:1
2:5
5:2
```

When the larger number is on the right, the colon function produces a vector
of numbers that increase by 1 each time you move to the right, until doing so
would exceed the larger number.  If the larger number is on the right, the
colon function does the same thing, but it decreases by 1 each time you move
an entry to the right.  It works this way even if the numbers you give are not
integers.

```{r}
2.5:5.2
5.2:2.5
```

If you want all of the entries of a vector with indices 3 through 7, you can
just use colon to ask for them:

```{r}
z <- c(2, 1, 5, 6, 3, 9, 1, 1); z
z[3:7]
```

R has other ways of slicing up vectors that you may find useful.  These include
the _head()_ and _tail()_ functions, as well as the use of negative indices.

### Naming the entries of a vector

When raw data is imported to R from an outside source, it is fairly common that
individual numbers are known by a name rather than an index value.  Due to this,
R has a fairly simple way of giving the entries of a vector descriptive names
in addition to numerical indices using the _names()_ function.  This function
is different than functions that we have seen so far because it is really
just a way of accessing a special property of your vector.  Be careful when you
combine named vectors, as the names may change in ways that you do not expect.



```{r}
x <- c(1, 5, 7)
names(x) <- c("A", "B", "C"); x

y <- c(1, 2, 5)
names(y) <- c("monkey", "1", "wow"); y

x + y
y + x
```

```{r}
x <- c(1, 5, 7)
names(x) <- c("A", "B", "C"); x
x + 1:3
x + 1:6
```

By default, R records that the names of the entries of your vector are NULL,
meaning that they do not exist.  If you name one of the entries, R will decide
that the others should be given names that are NA, meaning that they are
unknown.  If you set the names of your vector to NULL, then R goes back to
believing that the vector's entries should have no names.

```{r}
x <- c(1, 5, 7); x
names(x) <- c("A", "B", "C"); x
names(x) <- NULL;x
```

Once entries are named, you can pull them out by using those names just like
you did with indices

```{r}
x <- c(1, 5, 7); x
names(x) <- c("A", "B", "C"); x
x[c("C", "A")]
x[c(3, 1)]

```


