---
title: "04 - Matrices"
author: Michael Tychonievich, Ph.D.
output: html_notebook
---

### Matrices: R's default kind of "numerical array"

Often it is useful to have numerical data arranged in "spreadsheet" format
instead of as a linear list.  The most primitive way of doing this in R is by
recording the data in a matrix.  A matrix in R is essentially a vector with
extra information indicating how data is to be organized into rows and columns.
**CAUTION: All numbers in a single matrix must be of the same type!**  
If you place data of different types in the same matrix, then all entries will
be coerced into the most general data type present, just like with standard
vectors.  As with vectors, this
usually won't be a problem, but beware placing a character string in a vector
with numbers!  The simplest way to form a matrix from numbers that you happen
to have lying around is to use the _matrix()_ functionon a vector.  This
function is a bit more complicated to use than _c()_, so it's a good idea to
read some of R's internal documentation.

```{r}
?matrix
```

There are three different functions listed in the file, and the one you want
is the first one that creates a matrix from given values.  If you just use the
matrix function on a vector, it adds some structure to the vector.

```{r}
x <- c(2, 4, 6, 8, 10, 12, 14, 16, 18, 20)
M_0 <- matrix(x);M_0
```

Notice that the output here looks a lot like the vector _x_, but all of the
entries are listed in a column rather than in a row.  Over in the Environment
pane of RStudio, you can see how _M_0_ is different.  After _x_, you see that R
tells you that _x_ is "num [1:10]", to say that _x_ has a numeric class and
entries with indices 1 through 10.  After _M_0_, you see something similar: 
"num [1:10, 1]".  This tells you that _M_0_ is has a numeric class, and that it
is a rectangular array with 10 rows and 1 column.  When you make a matrix, you
can actually set these numbers!

Under usage, we are told what information the matrix function needs to form
its output.  All of the aruguments given are keyword arguments (kwargs)
with default values given.  To form a matrix from data that we have using
this function, we will need to know a few things: the data that we want to
put in the matrix (this will be a numerical vector, for now), either a number
of rows or columns to coerce the matrix to have, a decision about whether
"yrow should be TRUE or FALSE (this determines how the vector is laid out
in the matrix), and finally descriptive names for the row and column
dimensions if needed.  Let's make matrices from that vector that have different
shapes by specifying the value of the keywords nrow and ncol:

```{r}
x <- c(2, 4, 6, 8, 10, 12, 14, 16, 18, 20)
M_1 <- matrix(x, nrow=2);M_1
M_2 <- matrix(x, ncol=2);M_2
```

Notice that specifying either the number of rows or the number of columns is
enough to determine the matrix, so you need not specify both (and you should
not do it either, unless you have a good reason).  Look at what happens if try
to attach multiplie specifications:

```{r}
x <- c(2, 4, 6, 8, 10, 12, 14, 16, 18, 20)
M_3 <- matrix(x, ncol=2, nrow=2);M_3
M_4 <- matrix(x, nrow=2, ncol=2);M_4
```

By default, R wants to arrange your matrix so that the generating vector first
runs down the first column, then the second, and so on, until the matrix is
fully formed.  If you want to change this, you can use the byrow keyword

```{r}
x <- c(2, 4, 6, 8, 10, 12, 14, 16, 18, 20)
M_5 <- matrix(x, nrow=2, byrow=TRUE);M_5
M_6 <- matrix(x, ncol=2, byrow=TRUE);M_6
```

### Giving a matrix names

Just like with a vector, you can add names to a matrix.  There is a difference,
however: the names are associated with the rows and columns of the matrix
instead of the entries individually.  Let's make a named matrix using the
dimnames keyword argument.  The documentation for the _matrix()_ function says
that dimnames has to be a list of length 2, which means that a list has to be
defined!  You can do this with the _list()_ function.

```{r}
rownames = c("monkey", "tiger")
colnames = c("height", "weight")
dimnames = list(rownames, colnames)
animal <- matrix(c(2, 3, 60, 620), nrow=2, dimnames=dimnames)
animal
```

I did something a little bit strange there: I gave the list I wanted to use the
same name as the keyword that was supposed to represent it.  Despite this, R
didn't want to automatically grab my list and use it in the _matrix()_ function;
I had to explicitly tell it that my list was the correct one to use.  That's
what I did when when I wrote "dimnames=dimnames" as my keyword argument: I told
R to use the list that I named "dimnames" as the list used internally by the
_matrix()_ function, coincidentally also named "dimnames".  If I hand't done
that, then R would have used the _matrix()_ function's default internal value
for dimnames, which is NULL.

### Matrix "slicing"

You can now create matrices from data, but there will be times when you want to
get that data back out in pieces.  Because matrices have more structure than
vectors, it should make sense that there are more powerful ways of doing this.

First, you can get a single entry out of a matrix.  You can do this by using
indices or names.  When you ask for a single entry of a matrix, you should 
typically give both an indication of which row and also which column you want
the entry taken from.  It turns out that R still thinks of the matrix as being
a vector, to the point that you can still ask for and receive the entries of
the matrix by using a single index number.  Be careful with this!


```{r}
animal[1]
animal[1, 2]
animal["monkey", "weight"]
```

You can specify multiple names or indices by using vectors of indices if you
want to receive more information.  The output may not be formatted exactly how you would expect, however.  Tip: if you leave an index or name value blank, but you
make sure to leave the correct number of commas in your specification, R will
interpret the blank to mean "all possible values".  This can be a great way to
get entire rows or columns at once.

```{r}
animal[1, c(1, 2)]
animal[c("monkey", "tiger"), "weight"]
animal["tiger", ]
animal[ , "weight"]
animal[ , ]
```

### Tacking on rows or columns

With vectors, you could combine two vectors one after tha other using the _c()_
function.  An analogue of this for matrices is tacking on an extra row or
column to the matrix.  The functions that do these oeprations are _cbind()_ and _rbind()_.  Let's bind an extra column to the _animal_ matrix and give it a
name:

```{r}
animal <- cbind(animal, c(4,4))
colnames(animal)[3] <- "weight"
animal
animal[ , "weight"]
```

Oops!  I gave the new column the same name as one of the old ones, and now
I can't access the third column using its name.  Let's fix that.

```{r}
colnames(animal)[3] <- "limbs"
animal
```

Now let's use _rbind()_ to tack on a new row.  Since the matrix now has three
columns, this new row must have three entries.  This time, I will name the row
using _rbind()_ directly instead of _rownames()_ after the row is already added.
The syntax for this looks very similar to the syntax for using kwargs, so be
very careful if you decide to use this method.

```{r}
animal <- rbind(animal, beetle=c(0.4, 0.003, 6))
animal

```

And now it turns out that I used the wrong number for beetle weight.  To change
it, I first have to figure out how to access that entry of the matrix
individually.  Here are the three ways I showed you to do that:

```{r}
animal["beetle", "weight"]
animal[3, 2]
animal[6]

```

Now we just have to use one of these methods to put in a new value for the
weight of the beetle.

```{r}
animal["beetle", "weight"] = 0.0008
animal
animal[3, 2]
animal[6]

```

I used the row and column names to change the value in that entry, but you can
see that the other two methods of getting to the value in the beetle-weight
entry both give back the new value instead of the old one.

### Matrix arithmetic

Matrix arithmetic works similarly to vector arithmetic by default: arithmetic
operations are entry-by-entry by default.  If you want everything to go well,
make sure that the matrices you combine are the same size and expect that your
names might get mangled a bit.

```{r}
A <- matrix(1:4, nrow=2);A
dim(A)
B <- matrix(1:4, nrow=2, byrow=TRUE);B
dim(B)
A + B
A * B
A ^ B
A %% B
```

If you want to perform matrix multiplication or other operations that you saw
in your linear algebra class, you will need to use other syntax.  I won't
cover any of that here.

Exercises.

1.  What happens when you use the c() function to combine two matrices?

2.  What happens when you use the rep() function on a matrix?

3.  What happens when you use cbind() and rbind() with two matrices instead of
with a matrix and a vector?

4.  What happens when you add matrices of different sizes?

5.  Given numbers _n_ and _m_, how can I create an _n_ by _m_ matrix of all
zeros by using the matrix() function?