---
title: "05 - Logic"
author: Michael Tychonievich, Ph.D.
output: html_notebook
---

### Toward automation

When doing computations, there are a lot of operations that need to be done
repetitively.  To avoid this tedium, you are going to have to tell your
computer how it is to move from one task to the next another without waiting
for you to spell everything out.  At every moment, the computer will need to be
able to figure out what is happening now and use that information to decide
what to do next.

Let's start by looking at how we can ask the computer to get the information
it will need to make the decisions we tell it to make.

### Relations

R has a number of "information-gathering" functions already built in.  Some
of these functions yield up information about single objects, while others
look at how two or more objects relate to each other.  What these functions
all have in common is that they can be thought of as functions that answer
questions for which the only answers are something like "yes" and "no".
As implemented, each of these functions will always give back one of two values: 
TRUE or FALSE, the two objects of class logical.  Here is an example:

```{r}
x <- 5
is.logical(x)
```

In this example, I wanted to know something about the class of the number
stored as _x_.  Instead of asking R to tell me the class, I just asked it to
tell me if the class of x is "logical".  There are many properties that you can
access through _is_, but these properties may not be the ones you care about
initially.

### Numeric relations

The relations that you are probably used to dealing with in math class are
some special relations that compare the values or numbers, such as the
relations of "equals" and "less than".  R has special Boolean functions devoted
to just this sort of relation.

#### Equations

To ask if two objects are equal, you can use the following syntax:

```{r}
5 == 1 + 4
3 == 3 + 1
3 == "a"
"a" == "a"
```

Notice that I had to use two "equals" signs to signify that I wanted a logical
value.  If I had just put one equals sign, R would have thought I wanted to
assign a value to one of the variables!  This produces an error when you ask
to, say, assign the value 4 to 3, but it may not give you an error if your
equations involve variables!

To ask if two objects are *not* equal, use a single "equals"" sign preceded by an
exclamation point.  Think of this as an attempt to draw a line through the
"equals" sign.

```{r}
5 != 1 + 4
3 != 3 + 1
3 != "a"
"a" != "a"
```

#### Inequalities

When working with numbers, you will often want to say when one of a pair of
numbers is larger.  You can do this using the usual inequality symbols, 
"less than"(<)

```{r}
5 < 5
5 < 4
4 < 5
```

and "greater than (>)"

```{r}
5 > 5
5 > 4
4 > 5
```

As you just saw, these symbols are for strict inequalities.  For weak
inequalities, _i.e._ ones that should be true if the same number is given twice,
use the following syntax for "less than or equal to"

```{r}
5 <= 5
5 <= 4
4 <= 5
```

and "greater than or equal to"

```{r}
5 >= 5
5 >= 4
4 >= 5
```

The general rule to remember here is that the "equals" symbol is involved in
a comparison, it has to be the second one.

### Boolean functions

It will often be the case that you want to ask R to check on multiple
relations at the same time.  Perhaps you want something to happen when R
sees that the value assigned to x is between 0 and 1, but not 0.5.  For this
purpose, R has some functions built into it that act directly on logical values.
The four such functions are "and"(&), "or"(|), "not"(!), and "exclusive or"
(xor()).

```{r}

```


```{r}

```


```{r}

```
