---
title: "06 - Control: selection"
author: Michael Tychonievich, Ph.D.
output: html_notebook
---

### Automating decisions

When executing instructions, R's default behavior is to start at the top of a script and work its way downward, executing each line of code one after another. Most of the time this will be perfectly fine, but there will be times when you want R to skip over some code or to only execute certain instructions in special circumstances. For this we will need to use R's framework for what might be called "conditional execution", meaning execution only when certain conditions are satisfied.

### The structures `if`, `else if`, and `else`

The basic struction R has in place for conditional execution is the `if` statement. The idea is this: you give R a condition and some code. If R sees that the condition is `TRUE`, then it executes the code. Otherwise, R skips over the code and looks for an `else` statement. If R finds an `else`, then R executes the code associated with the else statement if it skipped the `if` code, and skips it otherwise. The `else` statement is optional! Try running the following code with different values of `x`, and note its behavior:

```{r}
x = 1
if (x < 0) {
  print("Negative.")
} else {
  print("Non-negative")
}
```

If it happens that you want R to check multiple conditions in sequence, executing code for the first condition that is found to be true, you can use the `else if` structure. Here is a more complicated sequence of conditions we can have R perform code based on:

```{r}
x <- 0.4

if (x < 0) {
  print("Negative.")
} else if (x == 0) {
  print("Zero.")
} else if (x > 0 & x < 1) {
  print("Between zero and one.")
} else if (x >= 1) {
  print("Greater than or equal to one.")
} else if (x < 10) {
  print("This should never be printed.")
} else {
  print("Something went very wrong.")
}
```

When R sees this block of code, it proceeds as follows. First, R evaluates the logical condition following the `if` statement. If it finds that the condition is `TRUE`, R executes that statement's code and jumps out of the `if`-`else if`-`else` block entirely. If R sees that the condition is false, it moves on to the next `else if` statement. If the condition given with this statement is `TRUE`, R executes the code following it and jumps to the end of the block. Otherwise, R moves on to the next `else if` and repeats the process. If no `else if` conditions evaluate to `TRUE`, then R finally executes the code following `else` (if present) and then continues on down the page.

This is why the code for the condition `x < 10` should never be executed in the above code: any number that you might give to this code should have already satisfied one of the earlier conditions, causing R to skip over this statement! This is one way to have R return a warning if some assumption that you made happens to not be true.

The effect of the `if`-`else if`-`else` structure is that we can tell R to choose which code to run from a list of possible blocks of code, using some condition that we have it compute on the fly.

You should note in this syntax that there are a lot of parentheses and braces use to separate different pieces. This punctuation is *actually* what tells R how to read your code; the line breaks are *optional*! But look at what happens when I decline to use line breaks and spaces in the conventional way:

```{r}
x<-0.5;if(x<0){print("Negative.")}else if(x==0){print("Zero.")}else if(x>0&x<1){print("Between zero and one.")}else if(x>=1){print("Greater than or equal to one.")}else{print("Something went wrong.")}
```

R has no problem reading this code, but most humans would find it somewhat challenging. It is squished together, everything is on the same line, and there is no "whitespace" around to set apart the different pieces of the code. It's a good idea to use line breaks and spaces in the conventional way if you ever want other people to be able to read your code. I count "yourself, next week" as "other people" for this purpose!

#### The `switch` structure

An alternative to `if` blocks in code is the use of a structure known as a `switch` statement. Rather than check general logical expression, a `switch` statement looks at a given character string variable and performs code based on what value it has. You can always use an `if` block instead of a `switch` block, but there are reasons (execution speed, code readability, personal style, etc.) to use it if you have that desire. I will avoid using it in these notebooks.

### `if`-`else` on a vector

By default, the `if` statement above does not work across vectors, instead only considering the first entry of a vector if asked to look at an entire vector at once. If you want an `if` statement to go through and check on every entry in a vector *in parallel*, the way to do that in R is to use the `ifelse()` function, which acts as a sort of "vectorized" `if` statement.

# Exercises.

1.  Use an if-else block to take a argument `x` that can be coerced to class `numeric` and print one of the following messages:

    * `"Not an integer."` if `x` is a number that is not an integer in the mathematical sense. In this sense, 1 is an integer despite the fact that `is.integer(1)` evaluates to `FALSE`.

    * `"Odd integer."` if `x` is an odd integer.

    * `"Even integer."` if `x` is an even integer.

    * `"Error!"` if `x` cannot be coerced to class `numeric`

2.  For the following vector `y`, return a vector `z` that has the same length, but, using the `ifelse()` function on `y`, make it so that:

    * every element that is greater than 1 is replaced by `1`

    * every element that is less than -1 is replaced by `-1`

    * every element that is between -0.3 and 0.3 is replaced by `0`

    * every other element is the same as the corresponding element of `y`

    Then generate a new value for y using the expression given for `x` and verify that your formula still works. Did you like this method better than using logical relations directly? Which one produces code that is more readable?

```{r}
x <- trunc(100*runif(10,-2,2))/100
y <- x
```

```{r}
z <- y 
x
z
```
