---
title: "06 - Control: selection"
author: Michael Tychonievich, Ph.D.
output: html_notebook
---

### Automating decisions

When executing instructions, R's default behavior is to start at the top of a
script and work its way downward, executing each line of code one after another.
Most of the time this will be perfectly fine, but there will be times when you
want R to skip over some code or to only execute certain instructions in
special circumstances.  For this we will need to use R's framework for what
might be called "conditional execution", meaning execution only when certain
conditions are satisfied.

### if-else if-else

The basic struction R has in place for conditional execution is the "if"
statement.  The idea is this: you give R a condition and some code.  If R sees
that the condition is TRUE, then it executes the code.  Otherwise, R skips over
the code and looks for an "else" statement.  If it finds an "else", then R
executes the code associated with the else statement if it skipped the "if"
code, and skips it otherwise.  Try running the following code with different
values of _x_, and note its behavior:

```{r}
x = 1
if (x < 0) {
  print("Negative.")
} else {
  print("Non-negative")
}
```

Note that the else statement here is actually entirely optional: there is no
requirement that you give R code to perform if your condition is not satisfied.

If it happens that you want R to check multiple conditions in a row, executing
code for which ever one is found to be true, you can use the "else if"
structure.  Here is a more complicated sequence of conditions we can have R
perform code based on:

```{r}
x <- 0.4

if (x < 0) {
  print("Negative.")
} else if (x == 0) {
  print("Zero.")
} else if (x > 0 & x < 1) {
  print("Between zero and one.")
} else if (x >= 1) {
  print("Greater than or equal to one.")
} else if (x < 10) {
  print("This should never be printed.")
} else {
  print("Something went very wrong.")
}
```

When R sees this block of code, it proceeds as follows.  First, R evaluates
each of the logical conditions following each if statement.  As soon as it
sees that one of them evaluates to TRUE, R executes that statement's code and
jumps out of the "if block".  If all of the conditions evaluate to FALSE, then
R executes the code after the else statement.  This is why the code for the
condition _x < 10_ should never be executed: any number that you might give
to this code should have already satisfied one of the earlier conditions!

The effect of this is that we can tell R to choose which code to run from a list
of possible blocks of code, using some condition that it computed.

You should note in this syntax that there are a lot of parentheses and braces
use to separate different pieces.  This punctuation is actually what tells R
how to read your code; the line breaks are the optional part.  Look at what
happens when I decline to use line breaks and spaces in a nice way:

```{r}
x <- 0.5
if(x<0){print("Negative.")}else if(x==0){print("Zero.")}else if(x>0&x<1){print(
"Between zero and one.")}else if(x>=1){print("Greater than or equal to one.")
}else{print("Something went wrong.")}
```
This code still functions, but it is very hard for a human to read.  It's
a good idea to use line breaks and spaces in a conventional way if you ever
want other people to be able to read your code.  I count "yourself, next week"
as "other people" for this purpose, by the way.

#### switch statements

An alternative to if blocks in code is the use of a structure known as a
switch statement.  Rather than check general logical expression, a switch
statement looks at a given character string variable and performs code based
on what value it has.
You can always use an if block instead of a switch statement, but there are
reasons (execution speed, code readability, personal style, etc.) to use it
if you have that desire.

### if-else on a vector

By default, the if statement above does not work across vectors, instead only
considering the first entry of a vector if asked to look at an entire vector
at once.  If you want an if statement to go through and check on every entry in
a vector in parallel, the way to do that in R is to use the _ifelse()_
function, which acts as a sort of "vectorized" if statement.

# Exercises.

1.  Use an if-else block to take a numeric argument _x_ and print one of the 
following messages:

* "Not an integer." if _x_ is a number that is not an integer in the
mathematical sense.  In this sense, 1 is an integer despite the fact that
is.integer(1) evaluates to FALSE.  You must exercise caution!

* "Odd integer." if _x_ is an odd integer.

* "Even integer." if _x_ is an even integer.

* "Error!" if _x_ fails to be a number.

2.  For the following vector _y_, return a vector that has the same length, but,
using the ifelse() function on _y_, make it so that:

* every element that is greater than 1 is replaced by 1

* every element that is less than -1 is replaced by -1

* every element that is between -0.3 and 0.3 is replaced by 0.

* every other element is the same as the corresponding element of _y_

Then generate a new value for y using the expression given for _x_ and verify
that your formula still works.
```{r}
x <- trunc(100*runif(10,-2,2))/100
```

```{r}
y <- x
```

Did you like this method better than using logical relations directly?  Which
one produces code that is more readable?