---
title: "07 - Control: loops"
author: Michael Tychonievich, Ph.D.
output: html_notebook
---

### More automation

Often in computations, the same operation needs to be repeated a number of times
to complete a problem.  In this situation, you are very fortunate if the same
computer code can be reused for each such iteration, since there is usually a
way to tell your computer to repeat the code until it is done.  In R, this
kind of repetition is done through loop structures.  There are three main loop
structures in R: the "for loop", the "while loop", and the "repeat loop".

### For loop

The safest form of loop to use is the for loop.  For a for loop, you must
first specify a vector or other list of values for R to iterate over.  R will
go through the list and perform your code one time for each entry on the list,
in the order that you appear on the list.  Note: don't alter this list within
the loop.  It won't do anything useful in the R language, and it sets up a bad
habit if you move on to certain other programming languages

Here is an example for loop that runs a simple block of code 10 times
to calculate compound interest.  For this code, I first some starting values,
then run the loop, and finally have R give me the output I wanted.

```{r}
principal <- 1000
interest <- 0.04
num_periods <- 10
amount <- principal

for (time in 1:num_periods) {
  amount <- amount * (1 + interest)
}

print(amount)
```

Since this code is more complicated than the code I have been giving you, I
have switched over to a convetion often used in programming: descriptive names.
When you see a quantity in a math problem, you usually want to give it a terse,
one-letter name like _x_.  This makes mathematical equations easier to read.

In this program, however, I have four different quantities to work with.  I
could give them short names, but that could spiral out of control very quickly
if I have to do a problem with a large number of inputs.  Fortunately,
the computer does not mind having to work with somewhat longer names, so I can
easily afford to tell it to use the entire word "principal" instead of just a
single "P".

A benefit to this convention is that now anyone looking at my code has a good
chance of understanding what my intentions were when I wrote the code, which
makes code maintenance less of a chore.

Now let's beef up that loop a little bit.  This time, I am going to have the
loop keep track of all of the intermediate balances before the final period
as well as the final amount.  I do this by changing _amount_ into a matrix and
then setting its entries to be the numbers I compute.

```{r}
# Initialization.
principal <- 1000
interest <- 0.04
num_periods <- 10
amount <- matrix(0, 1 + num_periods, 1)  # pre-allocation (optional)
amount[1] <- principal

# Compute values and place then in the matrix.
for (time in 1:num_periods) {
  amount[time + 1] <- amount[time] * (1 + interest)
}

# Display the results.
rownames(amount) <- c("start", 1:num_periods)
colnames(amount) <- "balance"
print(amount)
```

There is a lot going on in this code!  The most important thing to notice,
however, is that this loop will run a total of 10 times and no more, because
that is how long the vector used for iteration is.  To use a for loop, you
always have to know in advance how many times the loop is going to run, or at
least have a good upper bound.  This is what makes this sort of loop safer than
the other two kinds: by itself, it cannot run away from you and try to repeat
forever.

### Other loops

I will now go over the other two kinds of loops that you can use in R.  Unlike
for loops, these constructs can easily create endlessly running programs.
Please make sure to keep the following in mind:


**CAUTION: some of the following scripts may cause your interpreter to become**
**locked into an unending loop.  Either use the Esc key or Ctrl+C when the**
**console pane has focus to tell R to stop execution.**

#### while loops
In some circumstances it is useful to have an operation that repeats until
some condition is satisfied, instead of until a certain number of iterations
have been completed.  This is the situation in which a while loop is useful.

Suppose that in the prior problem setup, I really want to calculate how many 
time periods it will take to have 2,000 in my account instead.  Here is my plan:
I will tell R to keep iterating until the balance is over 2000.  Once that is
done, thenI will have R tell me the number of times it iterated.  I am pretty
sure that this will be done in a finite number of iterations, it's just that
I don't know how many until I do some kind of calculation (much of this could
be done by hand, too)

```{r}
principal <- 1000
interest <- 0.04
num_periods <- 0
amount <- principal

while (amount < 2000) {
  amount <- amount * (1 + interest)
  num_periods <- num_periods + 1
}

print(num_periods)
```

For a more interesting example, suppose that I am trying to pay off a loan.
With every time period (month) that passes, the loan accrues interest and then
I make a payment.  I can use a while loop to determine how long the loan will
take to pay off.

```{r}
num_payments <- 0
owed <- 200000.00
interest <- 0.003
payment <- 2000

while (owed > 0) {
  owed <- owed * (1 + interest)
  owed <- owed - payment
  num_payments <- num_payments + 1
}
final_payment = payment + owed
cat("I will make", num_payments - 1, "payments of", payment,
    "with one final payment of", final_payment, ". ",
    "Paying off this loan will take", num_payments/12, "years.\n")
```

To present the data that came out of this calculation, I had R print out a
summary message, using the _cat()_ function to turn a sequence of character
strings into a single string.

If you decide to change the numbers in this code, be very
careful!  If the monthly payment for the loan is too low, then the loan will
never be paid off and as such the code _will not stop on its own_.  A monthly
payment of 600 won't cause a problem, but going a little bit below that will.

#### Repeat

Repeat loops function as while loops, except that there is no condition that
makes a repeat loop stop.  That's right: repeat always makes infinite loops.
Does that seem a little bit strange to you?  Fortunately, there are ways to tell
a repeat loop that it is time to stop, which I will cover next.

### Loop-breaking statements

In a loop, you may decide that there are circumstances under which the loop
should stop early.  After all, if you have a for loop that is supposed to
iterate 100 times to look for something, and it finds it after 3 iterations,
then why would you want it to keep going?  R has two main ways to change the
flow of a loop: break and next.

#### break

When R sees a break statement, it immediately ends the last loop that it
entered (especially relevant if you put a loop within another loop).  It you
want to use repeat loops, the use of a break statement is mandatory if you want
the loop to ever stop.  The following code uses the repeat-break construct.  Try
to guess what it will do before you run it.

```{r}
N = 0
repeat {
  N <- N + 1
  print(N)
  if (N > 4) {
    break
  }
}

```

#### next

When R sees a next statement, it immediately ends the current iteration
of the last loop that it entered, and it moves on to the next iteration.  This
does not end the loop, but it does skip over any code in the loop that follows
the statement.  Try to determine what the following code will do before you run
it.

```{r}
N = 0
repeat {
  N <- N + 1
  if (N < 2) {
    next
  }
  print(N)
  if (N > 4) {
    break
  }
}

```

# Exercises.

1.  For the following vector _y_, return a vector that has the same length, but,
using the if-else blocks and a for loop, make it so that:

* every element that is greater than 1 is replaced by 1

* every element that is less than -1 is replaced by -1

* every element that is between -0.3 and 0.3 is replaced by 0.

* every other element is the same as the corresponding element of _y_

Then generate a new value for y using the expression given for _x_ and verify
that your formula still works.
```{r}
x <- trunc(100*runif(10,-2,2))/100
```

```{r}
y <- x
```

Did you like this method better than using logical relations directly?  How
about the ifelse() function?  Which one produces code that is most readable?

2.  Write a for loop that, given a vector, returns the value of the least
element.  Do not use min() or any other function in your script.

3.  Write a script that takes a numeric input _x_ and repeatedly cuts it in half
until it is smaller than 10^-10^, returning the result after the appropriate
number of halvings.  Do this in three different ways,

* Using a while loop

* Using a repeat loop with a break statement.

* Using a for loop with a break statement.  Make sure to loop over a vector that
is long enough, but don't try to make it "infinitely long."

* For a challenge, do this using no loops.  Logarithms will be helpful!