---
title: "08 - Functions"
author: Michael Tychonievich, Ph.D.
output: html_notebook
---

### Even more automation

The final method of controlling how a program executes that I will introduce is
the function.  A function in R is a sub-program within a script that you can
run any time after the interpreter has been told about it.  Though this may seem
like a small addition to your tool-kit, the ability to write functions gives
you a great deal of power!  Here is a simple example to get you started:

```{r}
f <- function(x) {
  y = x^2
  return(y)
}
f(3)
f(f(2))
```

Before explaining the syntax and how to use it, here is another function that
takes more than one argument and has a more descriptive name:

```{r}
multiply <- function(x, y) {
  z <- x * y
  return(z)
}
multiply(2, 4)
multiply(1, 0)
```

To define a function, first you write down the name that you want to give it.
You should name functions like you name variables: with descriptive and
memorable names of a "reasonable" length.  You assign to this name "function", 
followed by the list of inputs that you want to compute with in parentheses,
followed by the code for the function in braces, like this:

```{r}
good_name <- function(list_of_inputs) {what_the_function_does}
```

Whenever you place the appropriate number of inputs in parentheses following
your function name and run that code, R will execute the code you wrote for the
function with those values in place the the variable names that you used. An
important use for functions is, unsuprisingly, to execute the same code multiple
times in different places without having to write multiple copies of that code.

### Self-reference

With all of the power that the use of functions gives you, it's important to
realize that they are also quite dangerous.  Here is an example of a function
that involves recursion to reproduce the functionality of a while loop, with
all of the associated issues:

```{r}
danger <- function(x) {
  if (x < 5) {
    print("Danger")
    danger(x + 1)
  }
}

danger(0)
```

This sort of thing could easily run away on you, so be very cautious if you
decide to do it.  If you can manage, I recommend sticking with loops
so that others can more easily see your intentions.

### Global Environment and remove()

By now you may have noticed something about the functions that you've used so
far: they are all somehow living in the Global Environment, and you can see
references to them just below the list of variables you happen to have assigned.  This lets you know that you have continued access to those specific functions,
and you will continue to have access to them, even in other scripts and
notebooks that you might have open.  If you want to be rid of a function or
variable in your workspace without overwriting it, you can do so with the
following syntax:

```{r}
remove(f)
remove(good_name)
remove(multiply)
```

Even though R did not produce any output when those instructions were entered,
you should see that those functions are no longer present in your Global
Environment.  If you are wondering why you would ever want to do this, imagine
that you accidentally tell R that it should make pi exactly 3.

```{r}
pi
pi <- 3
pi
remove(pi)
pi
```

The remove() function actually restores the built-in definition for pi!

### Return()

A final note on function syntax: by default, R will return the last thing it
sees in a function computation as that function's value.  If you want to be
sure that the correct thing is being returned, then you can use return().  As
soon as R is asked to run a return() statement within a function, it will
stop execution of the function's code and return whatever is in the
parentheses as the function's output:

```{r}
f <- function(x) {
  return(x)
  print("No one will ever see this!")

}
f(3)
```

You can also use this as you would use a break statement in a loop to stop
the function from doing unneeded computations:

```{r}
f <- function(x) {
  if (x > 5) {
    return(x^2)
  }
  return(x^3)

}
f(2)
f(6)
```

### Scoping

Variables in R are divided up into workspaces.  There is a global workspace, and
it holds all of the variables that you define outside of function environments.
Each function you define also has its own local workspace, that only exists
while the function is being computed.  The local workspace holds any variable
that you passed into it by including it as one of your function's inputs, as
well as any variables that you define in the course of evaluating the function.
Variables that you pass in will still be around after the function is done
doing its work, at least as much as they were around before the function
was called upon.

If you define a variable within a
function's code, it will exist in this temporary local workspace for the
duration of the function's execution.  After the function is done, R's "garbage
collector" will notice (usually very quickly) that the quantity is no longer
being used, and the part of your computer's memory that held the quantity will
be cleared out.  If you use a name for a local variable that coincides with a
name for a global variable, this works just like how it worked when I
tried to change the value of pi: the new value is used until it is somehow
removed, either through the use of remove() or through the function ending
its execution:

```{r}
t <- 5
f <- function(x) {
  t <- 3
  print(t)
}
print(t)
f(0)
print(t)

```

Notice that the function printed its value of _t_, which was immediately
cleared out when the function was done, restoring the "old" value of t for the
final print().

As with global variables and built-in values, it is possible for a function to
see global variables without them being explicitly passed:

```{r}
t <- 5
f <- function(x) {
  print(t)
  t <- 3
  print(t)
}
f(0)
f(0)
```

This function knew the global value of _t_ and used it until it was given a new
value in the function's local workspace.  After the function was done, the local
value went away, and the global value again became the only one that R uses.
When the function executed the second time, it had to go back to the original
globally defined value!

If a function is defined within another function (a "local function"), something
similar happens: the local function has its own workspace, and it treats
it's containing function's workspace as a sort of global workspace.  But it can
still access the global workspace, just at a lower priority.  How functions
that are called in other functions access variables is determined by the
method that R uses for something called "scoping".  R happens to use a
method for scoping called "lexical scoping", which means that everything is
determined by _where the function was defined_ and not _where it is being used_.
If you want to use a function within another function and have it access that
function's workspace, it has to be a local function.

That said, the safest way to have one function access another function's
computations is by explicitly passing everything that you care about as function
arguments and never referring to variables within a function definition that the
function isn't given explicitly.  This approach makes your R functions behave
much more like mathematical functions: given the same inputs, they will always
return the same outputs.  This way of doing things is often known as 
"the functional programming paradigm."  It is not always possible to
adhere to this paradigm in R.
