---
title: "09 - Writing a Script"
author: Michael Tychonievich, Ph.D.
output: html_notebook
---

Let's revisit the mortgage payment example from the loops notebook:

```{r}
num_payments <- 0
owed <- 200000.00
interest <- 0.003
payment <- 2000

while (owed > 0) {
  owed <- owed * (1 + interest)
  owed <- owed - payment
  num_payments <- num_payments + 1
}
final_payment = payment + owed
cat("I will make", num_payments - 1, "payments of", payment,
    "with one final payment of", final_payment, ". ",
    "Paying off this loan will take", num_payments/12, "years.\n")
```

I would like to compare the time it takes to pay off the loan for many different
payment sizes, preferably with some kind of nice-looking printout.  I will start
by making the script into a function.  First, I will clear
out the global environment so that I will notice if I forget to account for any
of the many variables involved.

```{r}
remove(list=ls())
```

Now I will write the function.  It is going to need three inputs:
_owed_, _interest_, and _payment_.  I will go ahead and make these the arguments
of the function.  The variable *num_payments* will still be important in my
computation, but it looks more like a variable that the function itself must
determine.  That means that it should be a local variable instead of an input
argument.  Since my new goals do not involve having the report made in the
last step, I will also remove that.  

```{r}
payments <- function(payment=2000, interest=0.003, owed=200000){
  num_payments <- 0

while (owed > 0) {
  owed <- owed * (1 + interest)
  owed <- owed - payment
  num_payments <- num_payments + 1
}
return(num_payments)
}
payments()
payments(interest=.004)
payments(1500)
```

Notice that the function takes its arguements in the order
*(payment, interest, owed)*.  Deciding this was completely arbitrary on my part,
and that means that this function could be difficult to work with.  To improve
upon this, I made all of the input arguments into keyword arguments with some default values.  If
someone knows how I named the different quantities in my function, they can set
those quantities using my names instead of just relying on the order of inputs
to determine how arguments are set.  This is a definite break with how
mathematical notation works, but it can be very handy when lots and lots of
functions are floating around at the same time.  If no input is given for an
argument, then R will automatically put in my default value.

Now I would like to make a table of values for this function for different
values of the _payment_ input.  I do this with a for loop:

```{r}
num_trials <- 10
col_1 <- matrix(0, nrow=num_trials)
col_2 <- matrix(0, nrow=num_trials)

for (i in 1:num_trials) {
  payment = 1000 + 100*i
  col_1[i] = payment
  col_2[i] <- payments(payment=payment)  # This is weird!
}
T <- cbind(col_1, col_2)
colnames(T) <- c("payment",".003")
T
```

As I mentioned when I first covered kwargs, it sometimes happens that a global
variable and a keyword argument for a function share the same name, in this case
_payment_.  Because I set a default value for _payment_ within the payments()
function, the function will always ignore the global value associated with
_payment_ in favor of its own local value.  By writing "payment=payment" as an
arguement, I am telling the function directly to use the current global value of
_payment_ as the value it uses for its local variable, also named _payment_.
Despite their names being the same, these two variables can easily refer to
different things!

Now that I have a script to make a small table, I would like to increase the
scale once again.  This time I will make a function that repeatedly makes and
binds columns to the matrix I just made.  I will start this by reorganizing my
code a little bit.  The changes I make here will only affect how the code
carries out its computations: the output for a given list of inputs should
remain the same as the earlier code.  The process of doing this is often known
as "code refactoring".


```{r}
num_trials <- 10
col_1 <- matrix(1:10, nrow=num_trials)
col_1 <- 1000 + 100 * col_1

col_2 <- matrix(0, nrow=num_trials)

for (i in 1:num_trials) {
  col_2[i] <- payments(payment=col_1[i])
}
T <- cbind(col_1, col_2)
colnames(T) <- c("payment",".003")
T
```

This code does indeed give the same output as the earlier code, but it has an
advantage: the calculation of *col_1* happens outside of the loop for
calculating *col_2*, so I can more efficiently make new columns for my matrix.
I'll do this with a new function:

```{r}
num_trials <- 10
col_1 <- matrix(1:10, nrow=num_trials)
col_1 <- 1000 + 100 * col_1

col_maker <- function(interest, num_trials, payment_sizes) {
  col <- matrix(0, nrow=num_trials)
  for (i in 1:num_trials) {
    col[i] <- payments(payment=payment_sizes[i], interest)
  }
  return(col)
}

col_2 <- col_maker(.003, num_trials, col_1)

T <- cbind(col_1, col_2)
colnames(T) <- c("payment",".003")
T
```

Once again, this code gives me the same value for _T_, but now I have a simple
way to attach more columns.

```{r}
col_2 <- col_maker(.002, payment_sizes=col_1, num_trials=num_trials)
col_3 <- col_maker(.003, num_trials, col_1)
col_4 <- col_maker(.004, num_trials, col_1)
col_5 <- col_maker(.005, num_trials, col_1)

T <- cbind(col_1, col_2, col_3, col_4, col_5)
colnames(T) <- c("payment",".002",".003",".004",".005")
T
```

That is a much larger and more interesting table of values than what I had
before!  What's more, if I later use a formula to speed up the calculation of
the payments() function, I only have to mess around with that function to
upgrade the entire script.

Notice that R still lets me use keyword arguments in my function, even though I
did not give a default value to announce that I wanted that to happen.  Again,
this means that I can input arguments in a different order than what was
given when the function was defined.  Just to summarize, I will now place the
entire script below:

```{r}
remove(list=ls())

payments <- function(payment=2000, interest=0.003, owed=200000){
  num_payments <- 0
  while (owed > 0) {
    owed <- owed * (1 + interest) - payment
    num_payments <- num_payments + 1
  }
return(num_payments)
}

col_maker <- function(interest, num_trials, payment_sizes) {
  col <- matrix(0, nrow=num_trials)
  for (i in 1:num_trials) {
    col[i] <- payments(payment=payment_sizes[i], interest)
  }
  return(col)
}

num_trials <- 10
col_1 <- 1000 + 100 * matrix(1:10, nrow=num_trials)
col_2 <- col_maker(.002, num_trials, col_1)
col_3 <- col_maker(.003, num_trials, col_1)
col_4 <- col_maker(.004, num_trials, col_1)
col_5 <- col_maker(.005, num_trials, col_1)

T <- cbind(col_1, col_2, col_3, col_4, col_5)
colnames(T) <- c("payment",".002",".003",".004",".005")
T

```

I used functions to automate the construction of these data sets, and then
wrote out a list of very similar-looking commands to bind them up into a matrix
for viewing.  If I am satisfied with this code, this is a good place to stop.
But perhaps I would prefer to also automate the binding of columns.  That
would make the construction of a 10-column table a bit more bearable.  Let's
write a function to do this.  This time, I will use a vector of intest rates.

```{r}
num_trials <- 10
col_1 <- 1000 + 100 * matrix(1:10, nrow=num_trials)
make_payment_table <- function(interest_rates, num_trials, col_1) {
  T <- col_1
  colnames(T) <- "payment"
  for (rate in interest_rates) {
    new_col <- col_maker(rate, num_trials, col_1)
    colnames(new_col) <- rate
    T <- cbind(T, new_col)
  }
  return(T)
}
make_payment_table(c(.002,.003,.004,.005), num_trials, col_1)
```

That gets me my table with a single function call, so producing a large table
just requires me to change a few parameters:

```{r}
num_trials = 21
col_1 <- 100 * matrix(10:30, nrow=num_trials)
make_payment_table(.0005*(1:10), num_trials, col_1)
```

Here is the current state of the script:

```{r}
remove(list=ls())

payments <- function(payment=2000, interest=0.003, owed=200000){
  num_payments <- 0
  while (owed > 0) {
    owed <- owed * (1 + interest) - payment
    num_payments <- num_payments + 1
  }
return(num_payments)
}

col_maker <- function(interest, num_trials, payment_sizes) {
  col <- matrix(0, nrow=num_trials)
  for (i in 1:num_trials) {
    col[i] <- payments(payment=payment_sizes[i], interest)
  }
  return(col)
}

make_payment_table <- function(interest_rates, num_trials, col_1) {
  T <- col_1
  colnames(T) <- "payment"
  for (rate in interest_rates) {
    new_col <- col_maker(rate, num_trials, col_1)
    colnames(new_col) <- rate
    T <- cbind(T, new_col)
  }
  return(T)
}

num_trials = 21
col_1 <- 100 * matrix(10:30, nrow=num_trials)
make_payment_table(.0005*(1:10), num_trials, col_1)
```

This is a fairly complicated script at this point.  If I was to leave it like
this and come back tomorrow, I might forget what each of the functions I wrote
is supposed to work.  To help with that, I will alter the script by adding in
some comments.  I do this at the start of every function, and every time I do
something "clever."

```{r}
remove(list=ls())

payments <- function(payment=2000, interest=0.003, owed=200000){
  # Calculates the number of payments made on a loan with a given interest rate.
  # payment : the size of each loan payment
  # interest : the interest rate per period
  # owed : the initial loan amount
  num_payments <- 0
  while (owed > 0) {
    owed <- owed * (1 + interest) - payment
    num_payments <- num_payments + 1
  }
return(num_payments)
}

col_maker <- function(interest, num_trials, payment_sizes) {
  # creates a column of payment numbers in a table
  col <- matrix(0, nrow=num_trials)
  for (i in 1:num_trials) {
    col[i] <- payments(payment=payment_sizes[i], interest)
  }
  return(col)
}

make_payment_table <- function(interest_rates, num_trials, col_1) {
  # Binds together columns of payment numbers to make a table showing payment
  # numbers for various payment sizes and interest rates
  T <- col_1
  colnames(T) <- "payment"
  for (rate in interest_rates) {
    new_col <- col_maker(rate, num_trials, col_1)
    colnames(new_col) <- rate  # label new column with interest rate
    T <- cbind(T, new_col)
  }
  return(T)
}

num_trials = 21
col_1 <- 100 * matrix(10:30, nrow=num_trials)
make_payment_table(.0005*(1:10), num_trials, col_1)
```