---
title: "09 - Writing a Script"
author: "Michael Tychonievich, Ph.D."
output:
  html_document:
    df_print: paged
---

Let's revisit the mortgage payment example from the loops notebook.  I am to
expand upon the script we examined there to get it to generate some data that
I might be interested in.  Please note that most of what I will show you in this
notebook will be the style that I personally follow: you can certainly decide to
do things in a different way.


Here is the old script:

```{r}
num_payments <- 0
owed <- 200000.00
interest <- 0.003
payment <- 2000

while (owed > 0) {
  owed <- owed * (1 + interest)
  owed <- owed - payment
  num_payments <- num_payments + 1
}
final_payment = payment + owed
cat("I will make", num_payments - 1, "payments of", payment,
    "with one final payment of", final_payment, ". ",
    "Paying off this loan will take", num_payments/12, "years.\n")
```

I would like to compare the time it takes to pay off the loan for many different
payment sizes, preferably with some kind of nice-looking printout.  I will start
by making the script into a function.  First, I will clear
out the global environment so that I will notice if I forget to account for any
of the many variables involved.

```{r}
remove(list=ls())
```

Now I will write the function.  It is going to need three inputs:
_owed_, _interest_, and _payment_.  I will go ahead and make these the arguments
of the function.  The variable *num_payments* will still be important in my
computation, but it looks more like a variable that the function itself must
determine.  That means that it should be a local variable instead of an input
argument.  Since my new goals do not involve having the report made in the
last step, I will also remove that.  

```{r}
payments <- function(payment=2000, interest=0.003, owed=200000){
  num_payments <- 0

while (owed > 0) {
  owed <- owed * (1 + interest)
  owed <- owed - payment
  num_payments <- num_payments + 1
}
return(num_payments)
}
payments()
payments(interest=.004)
payments(1500)
```

Notice that the function takes its arguements in the order
*(payment, interest, owed)*.  Deciding this was completely arbitrary on my part,
and that means that this function could be difficult to work with.  To improve
upon this, I made all of the input arguments into keyword arguments with some default values.  If
someone knows how I named the different quantities in my function, they can set
those quantities using my names instead of just relying on the order of inputs
to determine how arguments are set.  This is a definite break with how
mathematical notation works, but it can be very handy when lots and lots of
functions are floating around at the same time.  If no input is given for an
argument, then R will automatically put in my default value.

Now I would like to make a table of values for this function for different
values of the _payment_ input.  I do this with a for loop:

```{r}
num_trials <- 10
col_1 <- matrix(0, nrow=num_trials)
col_2 <- matrix(0, nrow=num_trials)

for (i in 1:num_trials) {
  payment = 1000 + 100*i
  col_1[i] = payment
  col_2[i] <- payments(payment=payment)  # This is weird!
}
T <- cbind(col_1, col_2)
colnames(T) <- c("payment",".003")
T
```

As I mentioned when I first covered kwargs, it sometimes happens that a global
variable and a keyword argument for a function share the same name, in this case
_payment_.  Because I set a default value for _payment_ within the payments()
function, the function will always ignore the global value associated with
_payment_ in favor of its own local value.  By writing "payment=payment" as an
arguement, I am telling the function directly to use the current global value of
_payment_ as the value it uses for its local variable, also named _payment_.
Despite their names being the same, these two variables can easily refer to
different things!

Now that I have a script to make a small table, I would like to increase the
scale once again.  This time I will make a function that repeatedly makes and
binds columns to the matrix I just made.  I will start this by reorganizing my
code a little bit.  The changes I make here will only affect how the code
carries out its computations: the output for a given list of inputs should
remain the same as the earlier code.  The process of doing this is often known
as "code refactoring".


```{r}
num_trials <- 10
col_1 <- matrix(1:10, nrow=num_trials)
col_1 <- 1000 + 100 * col_1

col_2 <- matrix(0, nrow=num_trials)

for (i in 1:num_trials) {
  col_2[i] <- payments(payment=col_1[i])
}
T <- cbind(col_1, col_2)
colnames(T) <- c("payment",".003")
T
```

This code does indeed give the same output as the earlier code, but it has an
advantage: the calculation of *col_1* happens outside of the loop for
calculating *col_2*, so I can more efficiently make new columns for my matrix.
I'll do this with a new function:

```{r}
num_trials <- 10
col_1 <- matrix(1:10, nrow=num_trials)
col_1 <- 1000 + 100 * col_1

col_maker <- function(interest, num_trials, payment_sizes) {
  col <- matrix(0, nrow=num_trials)
  for (i in 1:num_trials) {
    col[i] <- payments(payment=payment_sizes[i], interest)
  }
  return(col)
}

col_2 <- col_maker(.003, num_trials, col_1)

T <- cbind(col_1, col_2)
colnames(T) <- c("payment",".003")
T
```

Once again, this code gives me the same value for _T_, but now I have a simple
way to attach more columns.

```{r}
col_2 <- col_maker(.002, payment_sizes=col_1, num_trials=num_trials)
col_3 <- col_maker(.003, num_trials, col_1)
col_4 <- col_maker(.004, num_trials, col_1)
col_5 <- col_maker(.005, num_trials, col_1)

T <- cbind(col_1, col_2, col_3, col_4, col_5)
colnames(T) <- c("payment",".002",".003",".004",".005")
T
```

That is a much larger and more interesting table of values than what I had
before!  What's more, if I later use a formula to speed up the calculation of
the payments() function, I only have to mess around with that function to
upgrade the entire script.

Notice that R still lets me use keyword arguments in my function, even though I
did not give a default value to announce that I wanted that to happen.  Again,
this means that I can input arguments in a different order than what was
given when the function was defined.  Just to summarize, I will now place the
entire script below:

```{r}
remove(list=ls())

payments <- function(payment=2000, interest=0.003, owed=200000){
  num_payments <- 0
  while (owed > 0) {
    owed <- owed * (1 + interest) - payment
    num_payments <- num_payments + 1
  }
return(num_payments)
}

col_maker <- function(interest, num_trials, payment_sizes) {
  col <- matrix(0, nrow=num_trials)
  for (i in 1:num_trials) {
    col[i] <- payments(payment=payment_sizes[i], interest)
  }
  return(col)
}

num_trials <- 10
col_1 <- 1000 + 100 * matrix(1:10, nrow=num_trials)
col_2 <- col_maker(.002, num_trials, col_1)
col_3 <- col_maker(.003, num_trials, col_1)
col_4 <- col_maker(.004, num_trials, col_1)
col_5 <- col_maker(.005, num_trials, col_1)

T <- cbind(col_1, col_2, col_3, col_4, col_5)
colnames(T) <- c("payment",".002",".003",".004",".005")
T

```

I used functions to automate the construction of these data sets, and then
wrote out a list of very similar-looking commands to bind them up into a matrix
for viewing.  If I am satisfied with this code, this is a good place to stop.
But perhaps I would prefer to also automate the binding of columns.  That
would make the construction of a 10-column table a bit more bearable.  Let's
write a function to do this.  This time, I will use a vector of intest rates.

```{r}
num_trials <- 10
col_1 <- 1000 + 100 * matrix(1:10, nrow=num_trials)
make_payment_table <- function(interest_rates, num_trials, col_1) {
  T <- col_1
  colnames(T) <- "payment"
  for (rate in interest_rates) {
    new_col <- col_maker(rate, num_trials, col_1)
    colnames(new_col) <- rate
    T <- cbind(T, new_col)
  }
  return(T)
}
make_payment_table(c(.002,.003,.004,.005), num_trials, col_1)
```

That gets me my table with a single function call, so producing a large table
just requires me to change a few parameters:

```{r}
num_trials = 21
col_1 <- 100 * matrix(10:30, nrow=num_trials)
make_payment_table(.0005*(1:10), num_trials, col_1)
```

Here is the current state of the script:

```{r}
remove(list=ls())

payments <- function(payment=2000, interest=0.003, owed=200000){
  num_payments <- 0
  while (owed > 0) {
    owed <- owed * (1 + interest) - payment
    num_payments <- num_payments + 1
  }
return(num_payments)
}

col_maker <- function(interest, num_trials, payment_sizes) {
  col <- matrix(0, nrow=num_trials)
  for (i in 1:num_trials) {
    col[i] <- payments(payment=payment_sizes[i], interest)
  }
  return(col)
}

make_payment_table <- function(interest_rates, num_trials, col_1) {
  T <- col_1
  colnames(T) <- "payment"
  for (rate in interest_rates) {
    new_col <- col_maker(rate, num_trials, col_1)
    colnames(new_col) <- rate
    T <- cbind(T, new_col)
  }
  return(T)
}

num_trials = 21
col_1 <- 100 * matrix(10:30, nrow=num_trials)
make_payment_table(.0005*(1:10), num_trials, col_1)
```

This is a fairly complicated script at this point: if I was to leave it like
this and come back tomorrow, I might forget what each of the functions I wrote
is supposed to do.  To help avoid that, I will alter the script by adding in
some comments.  I do this at the start of every function, and every time I do
something "clever."

```{r}
remove(list=ls())

payments <- function(payment=2000, interest=0.003, owed=200000){
  # Calculates the number of payments made on a loan with a given interest rate.
  # 
  # Inputs:
  #
  # payment : the size of each loan payment
  # interest : the interest rate per period
  # owed : the initial loan amount
  num_payments <- 0
  while (owed > 0) {
    owed <- owed * (1 + interest) - payment
    num_payments <- num_payments + 1
  }
return(num_payments)
}

col_maker <- function(interest, num_trials, payment_sizes) {
  # creates a column of payment numbers in a table
  # 
  # Inputs:
  #
  # interest : the interest rate per period
  # payment_sizes : list of payment sizes to check
  # num_trials : the number of payment sizes to check
  col <- matrix(0, nrow=num_trials)
  for (i in 1:num_trials) {
    col[i] <- payments(payment=payment_sizes[i], interest)
  }
  return(col)
}

make_payment_table <- function(interest_rates, num_trials, col_1) {
  # Binds together columns of payment numbers to make a table showing payment
  # numbers for various payment sizes and interest rates
  # 
  # Inputs:
  #
  # interest_rates : interest rates to check: determines the number of columns
  # num_trials : the number of payment sizes to check
  # col_1 : the column of payment sizes to check
  T <- col_1
  colnames(T) <- "payment"
  for (rate in interest_rates) {
    new_col <- col_maker(rate, num_trials, col_1)
    colnames(new_col) <- rate  # label new column with interest rate
    T <- cbind(T, new_col)
  }
  return(T)
}

num_trials = 21
col_1 <- 100 * matrix(10:30, nrow=num_trials)
make_payment_table(.0005*(1:10), num_trials, col_1)
```

That is much better, and as an added bonus other people who only get to see my
script code will now have a much better chance of knowing what I was thinking
when I wrote it!  This commenting also forced me to look back at my code.  There
are definitely some issues that I saw that I might want to deal with.  None of
these changes are required: after all, the program already does what I wanted
it to do!  Because of this, it's not a bad idea for me to write down all of
the _optional_ changes I might make to "improve"" it.  That way I can consider
which improvements are worth my time, and which should be neglected.  It would
also be a good time to set aside a backup copy of my code (if I haven't been
doing that all along) in case I manage to break something in a way that I don't
fully understand.  Here is my list of issues to consider:

* I used *col_1* as my name for the list of payment sizes I want to check in
some places, and *payment_sizes* in others.  Could I use just one of those
names?

* I keep demanding that the length of the various columns gets passed from
function to function, when I could have had each function recompute it using
the length() function.  Should I have done this instead?

* _T_ is not a very descriptive name for what I want out of that particular
compilation of data.  Should I change it?

* Sometimes I write my inputs using keywords and sometimes I don't.  Should I
stick with one approach over the other?

* I did not give the rows of the matrix _T_ any names.  Is that something I
Should consider?

* I happen to know that creating a new column and binding it to a matrix causes
R to make a new column and a new copy of the final matrix in my computer's
memory every time I do it.  This might make my program slower than it needs to
be; should I change the way new data is added to the matrix to avoid this?

* Through the use of functions, I have effectively put a while loop inside of a
for loop inside of another for loop.  This style of programming may prevent R
from performing "parallel processing", exploiting the fact that my computer has
multiple processor cores.  Should I try eliminate one of these loops?

* For certain combinations of interest rate and payment size, the loan under
consideration will never be paid off.  Should I make my program act differently,
perhaps returning a value of NA when that happens?  How can I tell if that is
going to happen?

* Should I put in default values for the arguments of more of my functions?

* The arguments that I used when I finally applied my functions at the end of
the script involve using :.  Is there a better way for me to have these
functions receive input?

As you can see, the list of issues I may want to address is already longer than
the script itself!  It would be straightforward to just go through this list and
make these changes, but it might be better for me in terms of productivity if I
set some priorities.  I am going to prioritize those issues that are either
quick to fix or could cause my program to break:

#### Quick fixes:

* Rename _T_ to *payment_table*

* Rename col_1 *to payment_sizes*

* Recompute length every time it is needed.

#### Possible game-breaker:

* Use a formula to calculate payment for a perpetuity, cut off calculations if
the payment size involves is larger than the payment of a perpetuity.
Alternatively, pick a big number and cut off anything that goes longer than
that.  Maybe 1200 periods?

#### Skippable:

* Skip optimization of loops since the program seems to run quickly enough.

* Leave the input format for make_payment_table() alone for now, and revisit if
it becomes an issue somehow.

```{r}
remove(list=ls())

# For a given list of interest rates and payment sizes, gives the number of
# periods required to pay off a mortgage loan of 200000.
#
# Functions:
# payments: for a given loan, returns the number of payments required
# column_maker: creates a column of payment numbers for various payment sizes
# make_payment_table: compiles those columns into a table for various rates

payments <- function(payment=2000, interest=0.003, owed=200000){
  # Calculates the number of payments made on a loan with a given interest rate.
  # 
  # Inputs:
  #
  # payment  : the size of each loan payment
  # interest : the interest rate per period
  # owed     : the initial loan amount
  num_payments <- 0
  # return NA if payment is less than the interest charged in the first period
  # or if the payment is "near" to that interest charge
  if (payment <= owed * interest || 
      isTRUE(all.equal(payment, owed * (1 + interest)))) {
    return(NA)
  }
  while (owed > 0) {
    owed <- owed * (1 + interest) - payment
    num_payments <- num_payments + 1
  }
return(num_payments)
}

column_maker <- function(interest=.003, payment_sizes=100*(10:20)) {
  # creates a column of payment numbers in a table
  # 
  # Inputs:
  #
  # interest : the interest rate per period
  # payment_sizes : list of payment sizes to check
  num_trials <- length(payment_sizes)
  col <- matrix(0, nrow=num_trials)
  for (i in 1:num_trials) {
    col[i] <- payments(payment=payment_sizes[i], interest=interest)
  }
  return(col)
}

make_payment_table <- function(interest_rates=.003, payment_sizes=100*(10:20)) {
  # Binds together columns of payment numbers to make a table showing payment
  # numbers for various payment sizes and interest rates
  # 
  # Inputs:
  #
  # interest_rates : interest rates to check: determines the number of columns
  # payment_sizes : the column of payment sizes to check
  payment_table <- payment_sizes
  colnames(payment_table) <- "payment"
  for (rate in interest_rates) {
    new_column <- column_maker(interest=rate, payment_sizes=payment_sizes)
    colnames(new_column) <- rate  # label new column with interest rate
    payment_table <- cbind(payment_table, new_column)
  }
  return(payment_table)
}

num_trials = 21
payment_sizes <- matrix(100*(10:30), nrow=num_trials)
make_payment_table(interest_rates=.0005*(1:10), payment_sizes=payment_sizes)
```

Now I feel like the script is done: I've got my table, the code is more
readable, and I prevented the situation where the program runs forever in a way
that isn't too disruptive, but that lets me know when it's triggered.  What
else might I do to improve the script?

# Exercises.

1.  Rewrite the final script by typing out everything _by hand_: no copying and
pasting.  You do not need to recopy the comments.  Go through your script and
change all of the descriptive variable names to names that are single letters.
Does this improve readability?  Check back in two weeks and ask yourself that
question again.

2.  Improve the script by turning it into a single function, making each of the
functions I wrote into local functions defined within its braces.  Make sure to
indent everything an extra time to make it visually apparent that you've done
this.

3.  Improve this script in the following way: take as initial inputs the
nominal interest rate per year and number of compoundings per year, and give 
the time to pay off the loan in units of years instead of periods.  Label each
column with the annual rate instead of the interest rate per period!  Change the
payment size so that equal payments are made each period, and label each row
with the payment size per "month."

4.  There is a formula to calculate the number of payments needed to pay off the
loan more directly.  Rewrite this script entirely to take advantage of this
formula.  You should be able to do this with fewer than three loops by using
vector and matrix operations.  It may help to review the different ways of
rounding numbers in R.